---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sitemapr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/bart-turczynski/sitemapr/actions/workflows/verify.yml/badge.svg)](https://github.com/bart-turczynski/sitemapr/actions/workflows/verify.yml)
[![Security audit](https://github.com/bart-turczynski/sitemapr/actions/workflows/security-audit.yml/badge.svg)](https://github.com/bart-turczynski/sitemapr/actions/workflows/security-audit.yml)
[![OSV audit](https://github.com/bart-turczynski/sitemapr/actions/workflows/osv-audit.yml/badge.svg)](https://github.com/bart-turczynski/sitemapr/actions/workflows/osv-audit.yml)
<!-- badges: end -->

sitemapr is a deterministic toolkit for reading and validating XML, text, and
index sitemaps against the [Sitemap Protocol
0.9](https://www.sitemaps.org/protocol.html) and the related W3C and RFC
standards. It parses sitemap sources into tidy tibbles, expands sitemap indexes,
discovers a site's sitemaps from its root URL, and reports schema and protocol
findings with stable, reproducible output.

> **Status: experimental.** The package works and is covered by an extensive
> test suite, but the API may still change without deprecation and it is not yet
> on CRAN. Pin a commit or tag if you depend on it.

## Installation

sitemapr and its URL-parsing dependency ([rurl](https://github.com/bart-turczynski/rurl))
are installed from GitHub. [pak](https://pak.r-lib.org) reads the `Remotes`
field and resolves the chain for you:

```r
# install.packages("pak")
pak::pak("bart-turczynski/sitemapr")
```

## Usage

### Read a sitemap into a tidy tibble

`read_sitemap()` accepts a sitemap URL or a local file (`.xml`, `.txt`, `.gz`,
or `.tar.gz`) and returns one row per URL, with `lastmod`, `changefreq`,
`priority`, and the image/video/news/alternate extension columns.

```{r read}
library(sitemapr)

sitemap <- '<?xml version="1.0" encoding="UTF-8"?>
<urlset xmlns="http://www.sitemaps.org/schemas/sitemap/0.9">
  <url>
    <loc>https://example.com/</loc>
    <lastmod>2024-01-01</lastmod>
    <priority>1.0</priority>
  </url>
  <url>
    <loc>https://example.com/about</loc>
    <changefreq>monthly</changefreq>
  </url>
</urlset>'

path <- tempfile(fileext = ".xml")
writeLines(sitemap, path)

read_sitemap(path)
```

A top-level sitemap index is expanded recursively (cycle-safe, depth- and
count-capped) so every reachable child sitemap's rows carry provenance.

### Validate against the schema and protocol

`validate_sitemap()` returns a stable findings tibble — one row per issue, with
a `code`, `severity`, and the `layer` (schema, protocol, classification, or
index) that produced it. The same source and mode always yield a row-for-row
identical result.

```{r validate}
invalid <- '<?xml version="1.0" encoding="UTF-8"?>
<urlset xmlns="http://www.sitemaps.org/schemas/sitemap/0.9">
  <url><loc>https://example.com/</loc><priority>2.0</priority></url>
  <url><loc>/relative/path</loc></url>
</urlset>'

path <- tempfile(fileext = ".xml")
writeLines(invalid, path)

findings <- validate_sitemap(path)
findings[, c("code", "severity", "layer")]
```

Pass `mode = "non-strict"` to downgrade schema violations to warnings and drop
strict-only findings.

### Discover a site's sitemaps

`sitemap_tree()` takes a site-root URL, tries a catalog of common sitemap paths
(generic, then WordPress/Shopify), and returns a discovery tree: one row per
candidate, marked `accepted` or `rejected`. It needs network access, so it is
not run here.

```{r tree, eval = FALSE}
sitemap_tree("https://example.com")
```

## Learn more

- Full reference and articles: <https://bart-turczynski.github.io/sitemapr/>
- Getting-started vignette: `vignette("introduction", package = "sitemapr")`

## Development

See [`CONTRIBUTING.md`](CONTRIBUTING.md) for the dev setup, verification gate,
and formatting conventions. Durable project context lives under
[`docs/`](docs/).
