---
title: "Introduction to sitemapr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sitemapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

sitemapr reads and validates XML, text, and index sitemaps against the
[Sitemap Protocol 0.9](https://www.sitemaps.org/protocol.html) and the related
W3C and RFC standards. Every function is deterministic: the same source yields
row-for-row identical output across calls, which makes results safe to snapshot
and diff.

This vignette walks through the three public entry points. The examples use
local files so they run without network access; each function also accepts a
URL.

```{r setup}
library(sitemapr)
```

## Reading a sitemap

`read_sitemap()` parses a source into a tidy tibble with one row per URL. It
accepts a URL or a local file (`.xml`, `.txt`, `.gz`, or `.tar.gz`) and returns
the `loc`, `lastmod`, `changefreq`, and `priority` columns alongside the
`images`/`video`/`news`/`alternates` extension list-columns and a
`source_sitemap` provenance column.

```{r read}
sitemap <- '<?xml version="1.0" encoding="UTF-8"?>
<urlset xmlns="http://www.sitemaps.org/schemas/sitemap/0.9">
  <url>
    <loc>https://example.com/</loc>
    <lastmod>2024-01-01</lastmod>
    <priority>1.0</priority>
  </url>
  <url>
    <loc>https://example.com/?page=about</loc>
    <changefreq>monthly</changefreq>
  </url>
</urlset>'

path <- tempfile(fileext = ".xml")
writeLines(sitemap, path)

urls <- read_sitemap(path)
urls[, c("loc", "lastmod", "changefreq", "priority")]
```

When the source is a sitemap **index**, `read_sitemap()` expands it recursively
— cycle-safe and capped on depth and child count — so every reachable child
sitemap's rows carry per-child provenance. The bounds are configurable through
the `index_limits` argument.

## Validating a sitemap

`read_sitemap()` never reports conformance problems; `validate_sitemap()` does.
It runs every finding producer — the XSD schema layer, the protocol/semantic
layer, the byte-level classification layer, and, for an index, the
index-expansion layer — and assembles the results into a stable findings
tibble.

```{r validate}
invalid <- '<?xml version="1.0" encoding="UTF-8"?>
<urlset xmlns="http://www.sitemaps.org/schemas/sitemap/0.9">
  <url><loc>https://example.com/</loc><priority>2.0</priority></url>
  <url><loc>/relative/path</loc></url>
</urlset>'

path <- tempfile(fileext = ".xml")
writeLines(invalid, path)

findings <- validate_sitemap(path)
findings[, c("code", "severity", "layer")]
```

Each finding carries a stable `code`, a `severity`, the `layer` that produced
it, and a `remediation_hint`, among other columns documented in the findings
contract. Two modes tune the strictness:

- `mode = "strict"` (the default) elevates the documented info-to-warning codes.
- `mode = "non-strict"` drops strict-only findings and downgrades schema
  violations to warnings.

A malformed source never crashes the call for a conformance reason: an HTML
page served where a sitemap was expected becomes an `UNSUPPORTED_HTML_MASQUERADE`
finding, and an unexpected XML root becomes an `UNSUPPORTED_ROOT` finding.
Only a genuine transport, SSRF, or HTTP failure raises a classed error.

## Discovering a site's sitemaps

Given a site-root URL, `sitemap_tree()` tries a catalog of common sitemap paths
(generic guesses first, then WordPress and Shopify conventions), classifies each
candidate as `accepted` or `rejected`, and returns a discovery tree — one row
per evaluated candidate, plus a row per expanded index child. A guessed path
that 404s is a `rejected` row, never an error, so a single unreachable guess
never fails the call.

```{r tree, eval = FALSE}
tree <- sitemap_tree("https://example.com")
tree[, c("depth", "sitemap_url", "status", "page_count")]
```

This example needs network access, so it is not evaluated here. robots.txt is
not consulted in this version.

## Where to go next

- The [function reference](https://bart-turczynski.github.io/sitemapr/reference/)
  documents every argument, including the `limits`, `index_limits`, and
  `user_agent` controls shared across the entry points.
- `read_sitemap()` for extraction, `validate_sitemap()` for conformance, and
  `sitemap_tree()` for discovery cover the full public surface.

## Acknowledgments

`sitemapr` implements the [Sitemaps.org](https://www.sitemaps.org/protocol.html)
protocol and Google's sitemap extensions, following the relevant W3C XML, RFC
3986/3987, and robots-exclusion standards. It is built on `xml2` (libxml2),
`httr2`, and the sibling `rurl` package.

The full list of credits — prior art, dependencies, the standards this code
implements, and the data sources it serves — is in
[`ACKNOWLEDGMENTS.md`](https://github.com/bart-turczynski/sitemapr/blob/main/ACKNOWLEDGMENTS.md).
